COFFEE: an Optimizing Compiler\\for Finite Element Local Assembly

### 2014-04-07 paper/bare_conf.tex
Incorporate Florian's suggestions

\section{Introduction}
In many fields, such as computational fluid dynamics, computational electromagnetics, and structural mechanics, phenomena are modeled by means of partial differential equations (PDEs). Numerical techniques, such as the finite volume and finite element method, are widely employed to approximate solutions of PDEs. Unstructured meshes are often used to discretize the computational domain, since their geometric flexibility allows adapting the mesh to the required resolution. The solution is sought in each element of the discretized domain by applying suitable numerical kernels. As the number of elements can be of the order of millions, the time required to execute the computation can be hours or days. To address this problem, domain-specific languages (DSLs) have been developed. The successful porting of Hydra, a computational fluid dynamics industrial application devised by Rolls Royce for turbomachinery design to OP2~\cite{pyop2isc} demonstrates the effectiveness of the DSL approach for implementing PDEs solvers~\cite{IstvanHydra}. Hydra uses the finite volume method, consists of roughly 50000 lines of code and is able to handle mesh sizes of over 100 millions edges.

% FIXME: that citation is wrong! You're talking about OP2. You MUST maintain that distinction

OP2 adopts a programming model, in which computations are expressed through self-contained functions, referred to as ``kernels''. A kernel is applied to all elements in a set of mesh entities such as edges, vertices, or cells, with an implicit synchronization between the application of two consecutive kernels. On commodity multi-cores, a kernel is executed sequentially by a thread, while parallelism is achieved partitioning the mesh and assigning each partition to a thread. Similar programming and execution models are adopted in~\cite{Fenics},~\cite{fluidity_manual_v4},~\cite{lizst}. Kernel optimization is one of the major concerns in unstructured mesh applications. In this paper, we tackle this problem by proposing an optimization strategy for a class of kernels used in finite element methods.

% FIXME: citation for the 30-60% figure?

We focus on local assembly (``assembly'' in the following), a fundamental step of the finite element method that covers an important fraction of the overall computation time, often in the range of 30$\%$-60$\%$. During the assembly phase, the solution of the PDE is approximated by executing a suitable kernel over all elements in the discretized domain. A kernel's working set is usually small enough to fit the L1 cache; it might need L2 cache when high-order methods are employed to improve the accuracy of the solution. However, we do not consider the latter case. An assembly kernel is characterized by the presence of an affine, often non-perfect loop nest, where individual loops are rather small: their trip count rarely exceeds 30, and may be as low as 3 for low-order methods. In the innermost loop, a problem-specific expression evaluates a two dimensional array, representing the result of local assembly in an element of the discretized domain. With such a kernel structure, we focus on aspects such as minimization of floating-point operations, register allocation and instruction-level parallelism, especially in the form of SIMD vectorization.

@@

Due to the constraints of available compilers and specialized linear-algebra libraries, we have automated a set of generic and model-driven code transformations in the COmpiler For FinitE Element local assembly (COFFEE), a compiler for optimizing local assembly kernels. COFFEE is integrated with Firedrake~\cite{firedrake-code}, a system for solving PDEs through the finite element method based on the PyOP2 abstraction~\cite{pyop2isc,pyop2ws}. It supports any problems expressible with this framework. This allows us to evaluate our code transformations in a range of real-world problems, varying key parameters that impact both solution accuracy and kernel cost, namely the polynomial order of the method (from $p=1$ to $p=4$) and the geometry of elements in the discretized domain (2D triangle, 3D tetrahedron, 3D prism).

% FIXME: You're talking about FFC-generated code really

Early experiments showed that Firedrake-generated code for non-trivial assembly kernels was sub-optimal. Our cost-model-driven sequence of source-to-source code transformations, aimed at improving SIMD vectorization and register data locality, can result in performance improvements up to 1.5$\times$ over ``softly-optimized'' code (that is code where only basic transformations are performed, such as generalized loop-invariant code motion, padding, and data alignment), and up to 4.44$\times$ over original kernels. The contributions of this paper are threefold:
\begin{itemize}
\item An optimization strategy for finite element local assembly, a class of kernels widely-used in scientific applications. Our approach exploits domain knowledge and goes beyond the limits of both vendor and research compilers.
\item Design and implementation of a compiler that automates the proposed code transformations for any problems expressible in Firedrake.
\item Systematic evaluation using a suite of examples of real-world importance, and evidence of significant performance improvements on two Intel architectures, a Sandy Bridge CPU and the Xeon Phi.
\end{itemize}

This paper is organized as follows: In Section~\ref{sec:background} we provide some background on local assembly, showing code generated by Firedrake and emphasizing important computational aspects. Section~\ref{sec:code-transf} describes the various code transformations in detail, highlighting when and how domain knowledge has been exploited. The design and implementation of our compiler is discussed in Section~\ref{sec:pyop2-compiler}. Section~\ref{sec:perf-results} shows performance results. Related work are illustrated in Section~\ref{sec:related-work}, while Section~\ref{sec:conclusions} concludes the paper.

@@

\begin{algorithm}[t]
\small
\KwIn{coordinate field (array), coefficient fields (e.g. velocity, array)}
\KwOut{element matrix A (2D array)}
\nlset{1}Compute Jacobian from coordinates \\
Declare of constant data: basis functions and derivatives\\
Compute element matrix using numerical quadrature
\caption{General structure of a local assembly kernel generated by Firedrake.}
\label{code:general-structure}
\end{algorithm}

@@

% FIXME: stiffness matrix is only used for certain types of elliptic PDEs
% FIXME: the fluidity manual isn't a good source for this, cite e.g. Brenner and Scott

Local assembly is the computation of local contributions of a specific cell of the discretized domain to the global sparse matrix or right-hand side vector of the linear system. The process consists of numerically evaluating an integral to produce a small dense element matrix or an element vector~\cite{quadrature1, fluidity_manual_v4}. In this work we focus on local matrices, which are usually more costly to compute.

% FIXME: you can't attribute the kernel code generation to Firedrake: mention FFC

Given a mathematical description of the input problem, expressed through the domain-specific Unified Form Language~\cite{ufl}, Firedrake generates C-code kernels implementing assembly using a numerical integration algorithm. These kernels are compiled at runtime using a vendor compiler, and Firedrake manages the parallel execution over the mesh. The subject of this paper is to enhance this execution model by adding an optimization stage prior to the generation of C code. The code transformations described next are also generalizable to non-Firedrake assembly kernels, provided that numerical integration is used.

The general structure of a Firedrake-generated kernel is shown in Figure~\ref{code:general-structure}. Inputs are a pointer to a zero initialized two dimensional array used to store the element matrix, a pointer to the element's coordinates in the discretized domain, and pointers to data of coefficient fields, for instance indicating value of velocity or pressure in the element. The output is the evaluated element matrix. A local assembly kernel can be logically split into three parts: 1) calculation of the Jacobian matrix, its determinant and its inverse; 2) definition of some constant two dimensional arrays for the finite element basis functions and their derivatives (if needed), initialized to problem-specific double-precision floating point values; 3) evaluation of the element matrix in an affine loop nest. All elements in the discretized domain share the same basis function arrays, so they are declared as global read-only arrays (using \texttt{static const} in C). Table~\ref{table:map-name-letters} shows the variable names we will use in the upcoming code snippets to refer to various objects in the kernel.

@@

% FIXME: You haven't really defined basis functions, so what is a Lagrange basis function?

The actual complexity of a local assembly kernel depends on the finite element problem being solved. In simpler cases, the loop nest is perfect: it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few products. An example is provided in Figure~\ref{code:helmholtz}, which shows an assembly kernel for a Helmholtz problem, using Lagrange basis functions on 2D elements with polynomial order $p=1$. In other scenarios, for instance when solving a non-linear problem like Burgers as shown in Figure~\ref{code:burgers}, the number of arrays involved in the computation of the element matrix can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression. Also, constants evaluated in outer loops (called $F$ in the code), acting as scaling factors of arrays, may be required; trip counts can be larger (proportionally to the order of the method); arrays may be block-sparse. Note that in addition to a larger number of operations to compute the element matrix, the Burgers case shows a register pressure higher than that in Helmholtz. Despite assembly kernels being problem-dependent, meaning that the space of codes that Firedrake can generate is infinite, it is still possible to identify common domain-specific traits that can be exploited for effective code transformations and SIMD vectorization.

Some characteristics of assembly kernels are: 1) memory accesses along the three loop dimensions are always stride-1; 2) the $j$ and $k$ loops are interchangeable, whereas permutation of $i$ might be subjected to pre-computation of values (e.g. the $F$ values in Burgers) and introduction of temporary arrays; 3) the $j$ and $k$ loops iterate over the same iteration space; 4) most of the sub-expressions on the right hand side of the element matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$). In Section~\ref{sec:code-transf} we show how to exploit these observations to define a set of systematic, composable optimizations.

\section{Code Transformations}
\label{sec:code-transf}
The code transformations presented in this section are applicable to all finite element problems that can be formulated in Firedrake. One particular characteristic is that they all aim at improving the run time of the \texttt{ijk} loop nest, where the numerical integration takes place. In rare, yet important cases, however, this might not be sufficient to achieve notable speed-ups. Generalized loop-invariant code motion, which is described in Section~\ref{sec:code-transf-licm}, is a fundamental optimization that allows pre-computation of invariant sub-expressions. It is worth noting that there are circumstances in which the amount of lifted terms, either at the level of an outer loop or even completely outside of the loop nest, is so big (thousands of operations, hundreds of temporaries) that their execution time becomes the dominant factor of the overall local assembly run-time. In these cases, two challenging optimizations are common sub-expression elimination and effective vectorization (superword level parallelism~\cite{SLP} would not be of help, because invariant sub-expressions lack, in general, structure). Given its inherent complexity, a comprehensive study of this problem is left as further work.

As already emphasized, the structure of mathematical expressions evaluating the element matrix and the variation in loop trip counts, although typically limited to the order of tens of iterations, render the optimization process challenging. While some problems achieve good performance relative to machine peak when applying a very specific set of transformations, there is another space of cases for which a more elaborate optimization strategy is required. For example, the Burgers problem in Figure~\ref{code:burgers}, given the large number of arrays accessed, suffers from high register pressure, whereas the Helmholtz problem in Figure~\ref{code:helmholtz} does not; this intuitively suggests that the two problems require a different treatment, based on an in-depth analysis of both data and iteration spaces. Furthermore, domain knowledge enables transformations that a general-purpose compiler could not apply, making the optimization space even larger. In this context, our goal is to understand the relationship between distinct code transformations, their impact on local assembly kernels, and to what extent their composability is effective in a class of problems and architectures.

% Padding and Data Alignment
\subsection{Padding and Data Alignment}
Auto-vectorization of assembly code computing the element matrix can be less effective if data are not aligned and if the length of the innermost loop is smaller than the vector length $vl$. Data alignment is enforced in two steps. Initially, all arrays are allocated to addresses that are multiples of $vl$. Then, two dimensional arrays are padded to a number of columns equal to the nearest multiple of $vl$. For instance, assume the original size of a basis function array is 3$\times$3 and that the underlying machine possesses AVX, which implies $vl=4$ since a vector register is 256 bits long and our kernels use 64-bits double-precision floating-point values. In this example, a padded version of the array will have size 3$\times$4. The compiler is explicitly informed about data alignment using a suitable pragma. Padding of all two dimensional arrays involved in the evaluation of the element matrix also allows us to safely round the loop trip count to the nearest multiple of $vl$. This avoids the introduction of a remainder (scalar) loop from the compiler, which would render vectorization less efficient.

% Autovectorizable LICM -> show code snippet?
\subsection{Generalized Loop-invariant Code Motion}
\label{sec:code-transf-licm}
From inspection of the codes in Figures~\ref{code:helmholtz} and~\ref{code:burgers}, it can be noticed that the computation of $A$ involves evaluating many sub-expressions that depend on two iteration variables only. Since symbols in most of these sub-expressions are read-only variables, there is ample space for loop-invariant code motion. Vendor compilers apply this technique, although not in the systematic way we need for our assembly kernels. We want to overcome two deficiencies that both Intel and GNU compilers have. First, they can identify sub-expressions that are invariant with respect to the innermost loop only. This is an issue for sub-expressions depending on $i$-$k$, which are not automatically lifted in the loop order \texttt{ijk}. Second, the hoisted code is scalar, i.e. it is not subjected to auto-vectorization. We work around these limitations with source-level loop-invariant code motion. In particular, we pre-compute all values that an invariant sub-expression assumes along its fastest varying dimension. This is implemented by introducing a temporary array per invariant sub-expression and by adding a new loop to the nest. At the price of extra memory for storing temporaries, the gain is that lifted terms can be auto-vectorized, since they are part of an inner loop. Given the short trip counts of our loops, it is important to achieve auto-vectorization of hoisted terms in order to minimize the percentage of scalar instructions, which could be otherwise significant. It is also worth noting that, in some problems, invariant sub-expressions along $j$ are identical to those along $k$ (e.g. in Helmholtz). In these cases, we safely avoid redundant pre-computation since, as anticipated in Section~\ref{sec:background}, a property of our domain is that $j$ and $k$ loops share the same iteration space.

@@

% Register allocation and register locality
\subsection{Model-driven Vector-register Tiling}
One notable problem of assembly kernels concerns register allocation and register locality. The critical situation occurs when loop trip counts and accessed variables are such that the vector-registers pressure is high. Since the kernel's working set fits the L1 cache, it is remarkably important to optimize register management. Canonical optimizations, such as loop interchange, unroll, and unroll-and-jam, can be employed to deal with this problem. In COFFEE, these optimizations are supported either by means of explicit code transformations (interchange, unroll-and-jam) or indirectly by delegation to the compiler through standard pragmas (unroll). Tiling at the level of vector registers is an additional feature of COFFEE. Based on the observation that the evaluation of the element matrix can be reduced to a ``summation of outer products'' along the $j$ and $k$ dimensions, a model-driven vector-register tiling strategy can be implemented. If we consider the code snippet in Figure~\ref{code:helmholtz-licm} and we ignore the presence of the operation \texttt{det*W3[i]}, we can notice that the computation of the element matrix is abstractly expressible as
\begin{equation}
\label{outer-product}
A_{jk} = \sum_{\substack{
  x \in B' \subseteq B \\
  y \in B'' \subseteq B}}
x_j\cdot y_k ~~~~~~ j,k = 0,...,4
\end{equation}
where $B$ is the set of all basis functions (or temporary variables, e.g. \texttt{LI$\_$0}) accessed in the kernel, whereas $B'$ and $B''$ are generic problem-dependent subsets. Regardless of the specific input problem, and by abstracting from the presence of all variables independent of both $j$ and $k$, the element matrix computation is always reducible to this form. Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 elements ($j,k=0,...,4$) of the element matrix using just 2 vector registers, which represent a 4$\times$4 tile, assuming $\vert B' \vert = \vert B'' \vert = 1$. Values in a register are shuffled each time a product is performed. Standard compiler auto-vectorization (GNU and Intel), instead, executes 4 broadcast operations (the ``splat'' of a value over all of the register locations) along the outer dimension to perform the calculation. Besides a larger number of cache accesses, it needs to keep between $f=1$ and $f=3$ extra registers to perform the same 16 evaluations when unroll-and-jam is used, with $f$ being the unroll-and-jam factor.

@@

The storage layout of $A$, however, is incorrect after the application of this outer-product-based vectorization (\emph{op-vect}, in the following). It can be efficiently restored with a sequence of vector shuffles following the pattern highlighted in Figure~\ref{fig:restore-layout}, executed once outside of the \texttt{ijk} loop nest. The generated pseudo-code for the simple Helmholtz problem when using \emph{op-vect} is shown in Figure~\ref{code:helmholtz-opvect}.

\begin{figure}[t]
\centerline{\includegraphics[scale=0.7]{Pictures/vect-restore.pdf}}
\caption{Restoring the storage layout after \emph{op-vect}. The figure shows how 4$\times$4 elements in the top-left block of the element matrix $A$ can be moved to their correct positions. Each rotation, represented by a group of three same-colored arrows, is implemented by a single vector shuffle intrinsics.}
\label{fig:restore-layout}
\end{figure}

@@

\begin{table}[h]
\begin{center}
\begin{tabulary}{1.0\columnwidth}{|C|C|}
\hline
Name (Abbreviation) & Parameter \\\hline\hline
Generalized loop-invariant code motion (\emph{licm}) &   \\ \hline
Padding &  \\ \hline
Data Alignment & \\ \hline
Loop interchange      & loops  \\ \hline
Loop unrolling  & unroll factor \\ \hline
Outer-product vectorization (\emph{op-vect}) & tile size \\ \hline
Expression splitting (\emph{split}) & split point, split factor \\ \hline
\end{tabulary}
\end{center}
\caption{Overview of code transformations for Firedrake-generated assembly kernels.}
\label{table:code-transformations}
\end{table}

@@

All loops are interchangeable provided that temporaries are introduced if the nest is not perfect. For the employed storage layout, the loop permutations \texttt{ijk} and \texttt{ikj} are likely to maximize performance. Conceptually, this is motivated by the fact that if the $i$ loop were in an inner position, then a significantly higher number of load instructions would be required every iteration; experiments showed that the performance loss is greater than the gain due to the possibility of accumulating increments in a register, rather than in memory, along the $i$ loop. The choice between \texttt{ijk} and \texttt{ikj} depends on the number of load instructions that can be hoisted out of the innermost dimension. Our compiler chooses, as outer, the loop along which the number of invariant loads is smaller so that more registers remain available to carry out the computation of the stiffness matrix.

>>>>

\section{Introduction}
In many fields, such as computational fluid dynamics, computational electromagnetics, and structural mechanics, phenomena are modelled by means of partial differential equations (PDEs). Numerical techniques, like finite volume method and finite element method, are widely-employed to approximate solutions of PDEs. Unstructured meshes are often used to discretize the equation domain, since their geometric flexibility allows solvers to be extremely effective. The solution is sought in each element of the discretized domain by applying suitable numerical operations. As the number of elements can be of the order of millions, a major issue is the time required to execute the computation, which can be hours or days. To address this problem, domain-specific languages (DSLs) have been developed. The successful porting of Hydra, a computational fluid dynamics industrial application devised by Rolls Royce for turbomachinery design (based on Finite Volume Method, roughly 50000 lines of code and mesh sizes that can be over 100 millions edges), to OP2~\cite{op2-main}, demonstrates the effectiveness of the DSL approach for implementing PDEs solvers~\cite{IstvanHydra}.

OP2 adopts a programming model in which computations are expressed through self-contained functions, referred to as ``kernels''. A kernel is applied to all elements in a set of mesh components, such as edges, vertices, or cells, with an implicit synchronization between the application of two consecutive kernels. On commodity multi-cores, a kernel is executed sequentially by a thread, while parallelism is achieved partitioning the mesh and assigning each partition to a thread. Similar programming and execution models are adopted in~\cite{Fenics},~\cite{fluidity_manual_v4},~\cite{lizst}. Kernel optimization is one of the major concerns in unstructured mesh applications. In this paper, we tackle this problem by proposing an optimization strategy for a class of kernels used in finite element methods.

We focus on local assembly (``assembly'', in the following), a fundamental step of a finite element method that covers an important fraction of the overall computation run-time, often in the range 30$\%$-60$\%$. During the assembly phase, the solution of the PDE is approximated by executing a suitable kernel over all elements in the discretized domain. A kernel's working set is usually small enough to fit the L1 cache; it might need L2 cache when high-order methods are employed to improve the accuracy of the solution. However, we do not consider the latter case. An assembly kernel is characterized by the presence of an affine, often non-perfect loop nest, where individual loops are rather small: their trip count rarely exceeds 30, and may be as low as 3 for lower order methods. In the innermost loop, a problem-specific expression evaluates a two dimensional array, representing the result of local assembly in an element of the discretized domain. With such a kernel structure, we focus on aspects like minimization of floating-point operations, register allocation and instruction-level parallelism, especially in the form of SIMD vectorization.

@@

Due to the constraints of available compilers and linear-algebra specialized libraries, we have automated a set of generic and model-driven code transformations in COFFEE\footnote{COFFEE stands for COmpiler For FinitE Element local assembly.}, a compiler for optimizing local assembly kernels. COFFEE is integrated with Firedrake~\cite{firedrake-code}, a system for solving PDEs through the finite element method based on the PyOP2 abstraction~\cite{pyop2isc}. It supports any problems expressible with this framework. This allows us to evaluate our code transformations in a range of real-world problems, varying key parameters that impact both solution accuracy and kernel cost, namely the polynomial order of the method (from $p=1$ to $p=4$) and the geometry of elements in the discretized domain (2D triangle, 3D tetrahedron, 3D prism).

Early experiments showed that Firedrake-generated code for non-trivial assembly kernels was sub-optimal. Our cost-model-driven sequence of source-to-source code transformations, aimed at improving SIMD vectorization and register data locality, can result in performance improvements up to 1.5$\times$ over ``softly-optimized'' code (i.e. code where only basic transformations are performed, such as generalized loop-invariant code motion, padding, and data alignment), and up to 4.44$\times$ over original kernels. The contributions of this paper are threefold
\begin{itemize}
\item An optimisation strategy for finite element local assembly. Our approach exploits domain knowledge and goes beyond the limits of both vendor and research compilers.
\item Design and implementation of a compiler that automates the proposed code transformations for any problems expressible in Firedrake.
\item Systematic evaluation using a suite of examples of real-world importance, and evidence of significant performance improvements on two Intel architectures, a Sandy Bridge CPU and the Xeon Phi.
\end{itemize}

The paper is organized as follows. In Section~\ref{sec:background} we provide some background on local assembly, showing code generated by Firedrake and emphasizing critical computational aspects. Section~\ref{sec:code-transf} describes the various code transformations, highlighting when and how domain-knowledge has been exploited. The design and implementation of our compiler is discussed in Section~\ref{sec:pyop2-compiler}. Section~\ref{sec:perf-results} shows performance results. Related work are illustrated in Section~\ref{sec:related-work}, while Section~\ref{sec:conclusions} concludes the paper.

@@

\begin{algorithm}[t]
\small
\KwIn{element matrix (2D array, initialized to 0),\\~~~~~~~~~element coordinates (array),\\~~~~~~~~~coefficient fields (array, e.g. velocity)}
\KwOut{element matrix (2D array)}
\nlset{1}~~- Compute Jacobian from coordinates \\
~~- Declare of constant data: basis functions and derivatives\\
~~- Compute element matrix using numerical quadrature
\caption{General structure of a local assembly kernel generated by Firedrake.}
\label{code:general-structure}
\end{algorithm}

@@

Local assembly is the computation of contributions of a specific cell in the discretized domain to the PDE solution. The process consists of numerically evaluating an integral to produce a small dense element matrix or an element vector~\cite{quadrature1, fluidity_manual_v4}. This operation is applied to all cells in the discretized domain. In this work we focus on local matrices, or ``element matrices'', which are more costly to compute.

Given a mathematical description of the input problem, expressed through the domain-specific Unified Form Language~\cite{ufl}, Firedrake generates C-code kernels implementing assembly using a numerical integration algorithm. It then triggers compilation of such kernels using an available vendor compiler, and eventually manages parallel execution over the mesh. The subject of this paper is to enhance this execution model by adding an optimization stage prior to the generation of C code. The code transformations described next are also generalizable to non-Firedrake assembly kernels, provided that numerical integration is used.

The general structure of a Firedrake-generated kernel is shown in Figure~\ref{code:general-structure}. Input are a pointer to a zero-initialized two dimensional array used to store the element matrix, a pointer to the element's coordinates in the discretized domain, and pointers to data of coefficient fields, for instance indicating value of velocity or pressure in the element. The output is the evaluated element matrix. A local assembly kernel can be logically split into three parts: 1) calculation of the Jacobian matrix, its determinant and its inverse; 2) definition of some constant two dimensional arrays, called ``basis functions'' (and their derivatives) in finite element terminology, initialized to problem-specific double-precision floating point values; 3) evaluation of the element matrix in an affine loop nest. All elements in the discretized domain share the same basis function arrays, so they are declared as global read-only arrays (i.e. using \texttt{static const} in C language). Table~\ref{table:map-name-letters} shows the variable names we will use in the upcoming code snippets to refer to the various kernel's objects.

@@

The actual complexity of a local assembly kernel depends on the finite element problem being solved. In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few products. An example is provided in Figure~\ref{code:helmholtz}, which shows an assembly kernel for a Helmholtz problem, using Lagrange basis functions on 2D elements with polynomial order $p=1$. In other scenarios, for instance when solving a non-linear problem, like Burgers in Figure~\ref{code:burgers}, the number of arrays involved in the computation of the element matrix can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression. Also, constants evaluated in outer loops (called $F$ in the code), acting as scaling factors of arrays, may be required; trip counts can be larger (proportionally to the order of the method); arrays may be block-sparse. Note that in addition to a larger number of operations to compute the element matrix, the Burgers case shows a register pressure higher than that in Helmholtz. Despite assembly kernels being problem-dependent, meaning that the space of codes that Firedrake can generate is infinite, it is still possible to identify common domain-specific traits that can be exploited for effective code transformations and SIMD vectorization.

Some peculiarities of assembly kernels are: 1) memory accesses along the three loop dimensions are always stride-1; 2) the $j$ and $k$ loops are interchangeable, whereas permutation of $i$ might be subjected to pre-computation of values (e.g. the $F$ values in Burgers) and introduction of temporary arrays; 3) the $j$ and $k$ loops iterate over the same iteration space; 4) most of the sub-expressions on the right hand side of the element matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$). In Section~\ref{sec:code-transf} we show how to exploit these observations to define a set of systematic, composable optimizations.

\section{Code Transformations}
\label{sec:code-transf}
The code transformations presented in this section are applicable to all finite element problems that can be formulated in Firedrake. One peculiar characteristic is that they all aim at improving the run-time of the \texttt{ijk} loop nest, where the numerical integration takes place. In rare, yet important cases, however, this might not be sufficient to achieve notable speed-ups. Generalized loop-invariant code motion, which is described in Section~\ref{sec:code-transf-licm}, is a fundamental optimization that allows pre-computation of invariant sub-expressions. It is worth noting there are circumstances in which the amount of lifted terms, either at the level of an outer loop or even completely outside of the loop nest, is so big (thousands of operations, hundreds of temporaries) that their execution time becomes the dominant factor of the overall local assembly run-time. In these cases, two challenging optimizations are common sub-expression elimination and effective vectorization (superword level parallelism~\cite{SLP} would not be of help, because invariant sub-expressions lack, in general, structure). Given its inherent complexity, a comprehensive study of this problem is left as further work.

As already emphasized, the structure of mathematical expressions evaluating the element matrix and the variation in loop trip counts, although typically limited to the order of tens iterations, render the optimization process challenging. Relatively to machine peak, while some problems achieve good performance when applying a very specific set of transformations, there is also an ample space of cases for which a more elaborated optimization strategy is required. For example, the Burgers problem in Figure~\ref{code:burgers}, given the large number of arrays accessed, suffers from high register pressure, whereas the Helmholtz problem in Figure~\ref{code:helmholtz} does not; this intuitively suggests that the two problems require a different treatment, based on an in-depth analysis of both data and iteration spaces. Furthermore, domain-knowledge enables transformations that a general-purpose compiler could not apply, making the optimization space even larger. In this context, our goal is to understand the relationship between distinct code transformations, their impact on local assembly kernels, and to what extent their composability is effective in a class of problems and architectures.

% Padding and Data Alignment
\subsection{Padding and Data Alignment}
Auto-vectorization of assembly code computing the element matrix can be less effective if data are not aligned and if the length of the innermost loop is smaller than the vector length $vl$. Data alignment is enforced in two steps. Initially, all arrays are allocated to addresses that are multiples of $vl$. Then, two dimensional arrays are padded by rounding the number of columns to the nearest multiple of $vl$. For instance, assume the original size of a basis function array is 3$\times$3 and that the underlying machine possesses AVX, which implies $vl=4$ since a vector register is 256 bits long and our kernels use 64-bits double-precision floating-point values. In this example, a padded version of the array will have size 3$\times$4. The compiler is explicitly informed about data alignment using a suitable pragma. Padding of all two dimensional arrays involved in the evaluation of the element matrix also allows us to safely round the loop trip count to the nearest multiple of $vl$. This avoids the introduction of a remainder (scalar) loop from the compiler, which would render vectorization less efficient.

% Autovectorizable LICM -> show code snippet?
\subsection{Generalized Loop-invariant Code Motion}
\label{sec:code-transf-licm}
From inspection of the codes in Figures~\ref{code:helmholtz} and~\ref{code:burgers}, it can be noticed that the computation of $A$ involves evaluating many sub-expressions that depend on two iteration variables only. Since symbols in most of these sub-expressions are read-only variables, there is ample space for loop-invariant code motion. Vendor compilers apply this technique, although not in the systematic way we need for our assembly kernels. We want to overcome two deficiencies that both $intel$ and $gnu$ compilers have. First, they can identify sub-expressions that are invariant with respect to the innermost loop only. This is an issue for sub-expressions depending on $i$-$k$, which are not automatically lifted in the loop order \texttt{ijk}. Second, the hoisted code is scalar, i.e. it is not subjected to auto-vectorization. We work around these limitations with source-level loop-invariant code motion. In particular, we pre-compute all values that an invariant sub-expression assumes along its fastest varying dimension. This is implemented by introducing a temporary array per invariant sub-expression and by adding a new loop to the nest. At the price of extra memory for storing temporaries, the gain is that lifted terms can be auto-vectorized, because part of an inner loop. Given the short trip counts of our loops, it is important to achieve auto-vectorization of hoisted terms in order to minimize the percentage of scalar instructions, which could be otherwise significant. It is also worth noting that, in some problems, invariant sub-expressions along $j$ are identical to those along $k$ (e.g. in Helmholtz). In these cases, we safely avoid redundant pre-computation since, as anticipated in Section~\ref{sec:background}, a property of our domain is that $j$ and $k$ loops share the same iteration space.

@@

% Register allocation and register locality
\subsection{Model-driven Vector-register Tiling}
One notable problem of assembly kernels concerns register allocation and register locality. The critical situation occurs when loop trip counts and accessed variables are such that the vector-registers pressure is high. Since the kernel's working set fits the L1 cache, it is remarkably important to optimize register management. Canonical optimizations, such as loop interchange, unroll, and unroll-and-jam, can be employed to deal with this problem. In COFFEE, these optimizations are supported either by means of explicit code transformations (interchange, unroll-and-jam) or indirectly by delegation to the compiler through standard pragmas (unroll). Tiling at the level of vector registers is an additional feature of COFFEE. Based on the observation that the evaluation of the element matrix can be reduced to a ``summation of outer products'' along the $j$ and $k$ dimensions, a model-driven vector-register tiling strategy can be implemented. If we consider the code snippet in Figure~\ref{code:helmholtz-licm} and we ignore the presence of the operation \texttt{det*W3[i]}, we can notice that the computation of the element matrix is abstractly expressible as
\begin{equation}
\label{outer-product}
A_{jk} = \sum_{\substack{
  x \in B' \subseteq B \\
  y \in B'' \subseteq B}}
x_j\cdot y_k ~~~~~~ j,k = 0,...,4
\end{equation}
where $B$ is the set of all basis functions (or temporary variables, e.g. \texttt{LI$\_$0}) accessed in the kernel, whereas $B'$ and $B''$ are generic problem-dependent subsets. Regardless of the specific input problem, and by abstracting from the presence of all variables independent of both $j$ and $k$, the element matrix computation is always reducible to this form. Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 entries ($j,k=0,...,4$) of the element matrix using just 2 vector registers, which represent a 4$\times$4 tile, assuming $\vert B' \vert = \vert B'' \vert = 1$. Values in a register are shuffled each time a product is performed. Standard compiler auto-vectorization ($gnu$ and $intel$), instead, executes 4 broadcast operations (i.e. ``splat'' of a value over all of the register locations) along the outer dimension to perform the calculation. Besides a larger number of cache accesses, it needs to keep between $f=1$ and $f=3$ extra registers to perform the same 16 evaluations when unroll-and-jam is used, with $f$ being the unroll-and-jam factor.

@@

The storage layout of $A$, however, is incorrect after the application of this outer-product-based vectorization ($op$-$vect$, in the following). It can be efficiently restored with a sequence of vector shuffles following the pattern highlighted in Figure~\ref{fig:restore-layout}, executed once outside of the \texttt{ijk} loop nest. The generated pseudo-code for the simple Helmholtz problem when using $op$-$vect$ is shown in Figure~\ref{code:helmholtz-opvect}.

\begin{figure}[t]
\centerline{\includegraphics[scale=0.7]{Pictures/vect-restore.pdf}}
\caption{Restoring the storage layout after $op$-$vect$. The figure shows how 4$\times$4 elements in the top-left block of the element matrix $A$ can be moved to their correct positions. Each rotation, represented by a group of three same-colored arrows, is implemented by a single vector shuffle intrinsics.}
\label{fig:restore-layout}
\end{figure}

@@

\begin{table}[h]
\begin{center}
\begin{tabulary}{1.0\columnwidth}{|C|C|}
\hline
Name (Abbreviation) & Parameter \\\hline\hline
Generalized loop-invariant code motion ($licm$) &   \\ \hline
Padding &  \\ \hline
Data Alignment & \\ \hline
Loop interchange      & loops  \\ \hline
Loop unrolling  & unroll factor \\ \hline
Outer-product vectorization ($op$-$vect$) & tile size \\ \hline
Expression splitting ($split$) & split point, split factor \\ \hline
\end{tabulary}
\end{center}
\caption{Overview of code transformations for Firedrake-generated assembly kernels.}
\label{table:code-transformations}
\end{table}

@@

All loops are interchangeable provided that temporaries are introduced if the nest is not perfect. For the employed storage layout, the loop permutations \texttt{ijk} and \texttt{ikj} are likely to maximize performance. Conceptually, this is motivated by the fact that if the $i$ loop were in an inner position, then a significantly higher number of load instructions would be required every iteration; experiments showed that the performance loss is greater than the gain due to the possibility of accumulating increments in a register, rather than in memory, along the $i$ loop. The choice between \texttt{ijk} and \texttt{ikj} depends on the number of load instructions that can be hoisted out of the innermost dimension. Our compiler chooses, as outer, the loop along which the number of invariant loads is smaller so that more registers remain available to carry out the computation of the element matrix.
